On the Functionality of the Multilayer Perceptron

### 2020-09-30 Thesis/Sections/FeatureSelections.tex

\paragraph*{}A spectrogram is produced by the method of \textit{frame-blocking}, which is very prevalent in audio signal classification. Frame-blocking creates a set of analysis frames, each labeled as $a^{(i)}$ and is $N$ samples in length, and has a fixed overlap with the next adjacent frame. Each of the $k$ frames then allows for a section of the signal in somewhat stationary state \cite{Liu,Zhang,Kahn,Virtanen}. For this project, we have chosen to use frames of size $N = 4096$ with a $75\%$ or $3072$ sample overlap. At a sample rate of $f_s = 44100$ samples/second, each frame represents a slice of time that is about $0.1$ seconds long.

\paragraph*{}We concatenate each analysis frame, $a^{(i)}, i \in [0,k-1]$ into a single $k \times N$ matrix, called $A$. Each row is a frame, each column is an index in that frame
\begin{equation}
\label{eqn-FrameMatrix}
A = \big\{ a^{(0)} , a^{(1)} , a^{(2)} , ... , a^{(k-1)} \big\} = 
\begin{bmatrix}
a^{(0)}[0] & a^{(0)}[1] & a^{(0)}[2] & \hdots & a^{(0)}[N-1] \\
a^{(1)}[0] & a^{(1)}[1] & a^{(1)}[2] & \hdots & a^{(1)}[N-1] \\
a^{(2)}[0] & a^{(2)}[1] & a^{(2)}[2] & \hdots & a^{(2)}[N-1] \\
\vdots 		& \vdots 	  & \vdots 		& \ddots & \vdots        \\
a^{(k-1)}[0] & a^{(k-1)}[1] & a^{(k-1)}[2] & \hdots & a^{(k-1)}[N-1] \\
\end{bmatrix}
\end{equation}
We use bracket notation, $[j]$ to indicate that each object is array-like. Many programming languages could also use the following indexing conventions for matrix 
$A$:
\begin{equation}
\label{eqn-IndexingA}
A_{i,j} = a^{(i)}[j] = A[i][j] = A[i,j]
\end{equation}
Note that these all represent equivalent entries.

\paragraph*{}After frame-blocking, we apply a \textit{windowing function} to each frame. A standard \textit{Hann Window} of $N$ samples is generated and reshaped into a $N \times 1$ column-array, $H$. The $n$-th index in a Hann window with $N$ samples is defined:
\begin{equation}
\label{eqn-Hann}
H[n] = \frac{1}{2}\bigg[ 1 - \cos\Big( \frac{2\pi n}{N-1}\Big)\bigg]
\end{equation}
The window function is applied to each analysis frame by computing the dot-product of the Hann window array, $H$ and each row of the analysis frames matrix, $A_{i,:}$. Result is an $k \times N$ array, $\widetilde{A}$:
\begin{equation}
\label{eqn-WindowMatrix}
\widetilde{A} = A \cdot H^T
\end{equation}

@@

\paragraph*{}Most standard implementations of neural network models require all activations, weights, and biases to be real floating-point numbers. Since the the DFT matrix introduces complex values, we compute the square of the element-wise $L2$-norm of the resultant array.
\begin{equation}
\label{eqn-Spectrogram}
S = \big{\lVert} \mathbb{W}\widetilde{A}^T \big{\rVert}_{2}^{2}
\end{equation}
Where $\mathbb{W}$ is the DFT matrix from eqn. (\ref{eqn-DFTMatrix}) and $\widetilde{A}^T$ is the transpose of the analysis frames matrix from eqn. (\ref{eqn-WindowMatrix}). The matrix $S$, is the spectrogram representation of the initial waveform, and has shape $N \times k$ of real floating-point numbers. We can index matrix $S$ similarly to that of matrix $A$ in eqn. (\ref{eqn-IndexingA}):
\begin{equation}
\label{eqn-IndexingS}
S_{i,j} = s^{(i)}[j] = S[i][j] = S[i,j]
\end{equation}

@@

\paragraph*{}However, standard musical instruments seldom extend above $8$ kHz \cite{Olson,Virtanen,White}. This means that when constructing the spectrogram, we will rarely ever see energy present above this frequency at any time, and the $S$ matrix will contain mostly zero, or zero-like entries. As a result of this, we can simply crop each matrix, to display frequencies that are only between $0$ Hz and $8000$ Hz. This makes the input array smaller, and eliminates redundant and non-useful information. The number rows in the $S$ matrix is reduced from $N$ down to $N'$. lastly, each raw audio file may contain a different number of samples to begin with, thus a different number of analysis frames are created. To ensure all samples are a homogeneous size, we assert that each matrix must have exactly $k'$ columns. If $k' < k$, columns of zeros are added to pad the input, and if $k' > k$, columns are removed.

@@

\paragraph*{}The features described in this section are derived from time-domain representations of each audio sample. In some cases, this can be from the waveform directly, or my manipulation of rows and columns in the analysis frames matrix $A$ from eqn. (\ref{eqn-FrameMatrix}). For each feature, we detail the physical significance and provide a visualization in feature-space.

\paragraph*{}From time space, we use the following $7$ features:

\begin{itemize}

\item[•] Time Domain Envelope

\item[•] Zero Crossing Rate

\item[•] Temporal Center of Mass

\item[•] Auto Correlation Coefficients ($\times 4$)

\end{itemize}

@@

\paragraph*{}The time domain envelope (TDE) is a method of determining which parts of the sound wave contains most of the energy of the signal. Generally, we start by computing the RMS-Energy of each frame. The larger the RMS energy of the frame, the larger the amplitude of the waveform in that frame. For frames with smaller amplitude, we conclude that the signal has either not begun, or is decaying. The RMS-Energy of a single analysis-frame $a^{(i)}$ is given by \cite{Olson,Virtanen}
\begin{equation}
\label{eqn-RMS}
\text{RMS}\big[ a^{(i)} \big] = \sqrt{\frac{1}{N} \sum_{j=0}^{N-1} \big( a^{(i)}[j]} \big)^2
\end{equation}

\paragraph*{}We adapt this feature to \textbf{compute the RMS-Energy of the full waveform}. By doing this, we acquire a rough estimate for the energy of a waveform in the entirety of the audio file \cite{Liu}. For instruments with long sustain or release times, such as strings or undampened mallet percussion, hold a comparably large waveform RMS when compared to those instruments without sustain such as brass or percussion.

@@

\paragraph*{}The zero crossing rate (ZXR) of a signal or frame is use to measure how many times that a signal crosses it's equilibrium point. This can be done per total sound wave, or per unit time, such an per analysis-frame. In differentiating speech and music, it is commonly to use the difference between the frame with the highest ZXR and the lowest ZXR. In some cases, the ZXR can be correlated to frequency as well \cite{Kahn,Zhang}. The ZXR for a frame $a^{(i)}$ is given by \cite{Virtanen,Liu}
\begin{equation}
\label{eqn-ZXR}
\text{ZXR}\big[ a^{(i)} \big] = \frac{1}{2} \sum_{j=1}^{N-1} \big{|} \sign(a^{(i)}[j]) - \sign(a^{(i)}[j-1]) \big{|} 
\end{equation}
Where $\sign(x)$ returns $+1$ if $x > 0$, $-1$ if $x < 0$ and $0$ if $x = 0$.

\paragraph*{}We adapt this feature to \textbf{compute the zero crossing rate for the full waveform}. This provides a rough estimate for the average frequency in the full waveform, and can help discern clean periodic signals (low ZXR) from those that may have more volatile behavior (high ZXR)\cite{Virtanen}.

@@

\paragraph*{}The temporal center of mass (TCM) of a signal is used to compute where in time the amplitude sort of \textit{bunches up}. We treat the full waveform array, $a$, with $M$ samples as a 1-Dimensional discrete mass distribution. The TCM of that waveform is then given:
\begin{equation}
\label{eqn-FeatureTCM}
\text{TCM}\big[ a \big] = \frac{\sum_{j=0}^{M-1}j a[j]}{\sum_{j=0}^{M-1}a[j]}
\end{equation}
This includes negative amplitude values acting as "negative masses".

\paragraph*{}The TCM turns out to be a very unique feature, and very powerful in classification due to it's variance. For instruments with heavier attacks, we expect the TCM to be a very low value as most of the energy and amplitude is concentrated close to the start of the audio file. For instruments with long sustain times, we expect a very centrally located center of mass, and for instruments with long release times, we expect a very late TCM value.

@@

\paragraph*{}Auto correlation coefficients (ACC) are rough estimates of the signal spectral distribution \cite{Virtanen}. We can compute any number of ACC's and their value changed depending on the index chosen. It is common to only compute the first $k$ ACC's. For a full waveform signal $a$, with $M$ samples, the $k$-th ACC (indexed from $1$ to $K$) is given by:
\begin{equation}
\label{eqn-FeatureACC}
\text{ACC}_k\big[ a \big] = \frac{\sum_{j=0}^{M-k-1}a[j]a[j+k]} {\sqrt{\sum_{j=0}^{M-k-1}a^2[j]} \sqrt{\sum_{j=0}^{M-k-1}a^2[j+k]}}
\end{equation}

@@

\paragraph*{}The features described in this section are derived from the frequency-domain representations of each audio sample. Each of these features are exacted from the transpose spectrogram representation, $S^T$, where the creation of $S$ is outlined in sec. (\ref{subsec-spectrogram}) and computed in eqn.(\ref{eqn-Spectrogram}). This is done to return analysis frame into a row-like array, representing frequency space divided into $N'$ bins, with $k'$ analysis frames. For each feature, we detail the physical significance and provide a visualization in feature-space.

\paragraph*{}From frequency space, we use the following $13$ features:
\begin{itemize}

\item[•] Mel Frequency Cepstral Coefficients ($\times 12$)

\item[•] Frequency Center of Mass

\end{itemize}

@@

\paragraph*{}Mel Frequency Ceptstral Coefficients (MFCC's) are a very prolific feature used widly in audio classification tasks \cite{Liu,Virtanen,Zhang}. MFCC's are produced by converting a linear-frequency axis with units such as Hertz into units of Mels, which are design to account for the non-linear perception of frequency in humans. The Hertz to Mel and Mel to Hertz transforms are provided \cite{Virtanen}:
\begin{equation}
\label{eqn-HztoMel}
M_f[h] = 2595 \log_{10}\big(1+ \frac{h}{700}\big)
\end{equation}
\begin{equation}
\label{eqn-MeltoHz}
H_f[m] = 700 \big(10^{(\frac{m}{2595})}-1\big)
\end{equation}
Where $M_f$ is the frequency in units of Mels, given $[h]$, a frequency in Hertz, and $H_f$ is the frequency in Hertz given $[m]$ a frequency in Mels.

@@

\paragraph*{}The frequency center-of-mass (FCM) for an audio file provides a strong representation of how overtones and energy is distributed in the signal's frequency domain. As with the temporal center-of-mass, we treat each row of the $S$ matrix as it's own 1-Dimensional mass distribution, and compute the center of mass of that row. This encodes the FCM for a single analysis frame (in frequency-space) in the waveform. For an frequency-analysis frame, $s^{(i)}$, the FCM is given by:
\begin{equation}
\label{eqn-FeatureFCM}
\text{TCM}_{i}\big[ s^{(i)} \big] = \frac{\sum_{j=0}^{N'-1}j s^{(i)}[j]}{\sum_{j=0}^{N'-1}s^{(i)}[j]}
\end{equation}
We compute the FCM for each of the $k'$ frames, and then average the results. We use the average FCM across $k'$ frames to compute the FCM feature:
\begin{equation}
TCM = \sum_{i=0}^{k'} \text{TCM}_{i}\big[ s^{(i)} \big]
\end{equation}

>>>>

\paragraph*{}A spectrogram is produced by the method of \textit{frame-blocking}, which is very prevalent in audio signal classification. Frame-blocking takes a raw waveform or signal, $s$ and decomposes it into a set of analysis frames, $a_i$, with each being $N$ samples in length, and has a fixed overlap with the next adjacent frame. Each of the $k$ frames then allows for a section of the signal in somewhat stationary state \cite{Liu,Zhang,Kahn,Virtanen}. For this project, we have chosen to use frames of size $N = 4096$ with a $75\%$ or $3072$ sample overlap. At a sample rate of $f_s = 44100$ samples/second, each frame represents a slice of time that is about $0.1$ seconds long.

\paragraph*{}We concatenate each analysis frame, $a_{i}, i \in [0,k-1]$ into a single $k \times N$ matrix, called $A$. Each row is a frame, each column is an index in that frame
\begin{equation}
\label{eqn-FrameMatrix}
A = \big\{ a_{0} , a_{1} , a_{2} , ... , a_{k-1} \big\} = 
\begin{bmatrix}
a_{0}[0] & a_{0}[1] & a_{0}[2] & \hdots & a_{0}[N-1] \\
a_{1}[0] & a_{1}[1] & a_{1}[2] & \hdots & a_{1}[N-1] \\
a_{2}[0] & a_{2}[1] & a_{2}[2] & \hdots & a_{2}[N-1] \\
\vdots 		& \vdots 	  & \vdots 		& \ddots & \vdots        \\
a_{k-1}[0] & a_{k-1}[1] & a_{k-1}[2] & \hdots & a_{k-1}[N-1] \\
\end{bmatrix}
\end{equation}
We use bracket notation, $[j]$ to indicate that each analysis frame $a_i$ is array-like. Many programming languages could also use the following indexing conventions for matrix $A$:
\begin{equation}
\label{eqn-IndexingA}
A_{i,j} = A_i[j] = A[i][j] = A[i,j]
\end{equation}
These all represent equivalent entries.

\paragraph*{}After frame-blocking, we apply a \textit{windowing function} to each frame. A standard \textit{Hann Window} of $N$ samples is generated and reshaped into a $N \times 1$ column-array, $H$. The $n$-th index in a Hann window with $N$ samples is defined:
\begin{equation}
\label{eqn-Hann}
H[n] = \frac{1}{2}\bigg[ 1 - \cos\Big( \frac{2\pi n}{N-1}\Big)\bigg]
\end{equation}
The window function is applied to each analysis frame by computing the element-wise product of the Hann window array, $H$ and each row of the analysis frames matrix, $A_{i,:}$. Result is an $k \times N$ array, $\widetilde{A}$:
\begin{equation}
\label{eqn-WindowMatrix}
\widetilde{A_i} = A_i \odot H
\end{equation}

@@

\paragraph*{}Most standard implementations of neural network models require all activations, weights, and biases to be real floating-point numbers. Since the the DFT matrix introduces complex values, we compute the square of the element-wise $L2$-norm of the resultant array.
\begin{equation}
\label{eqn-Spectrogram}
S_{i,j} = \big{\lVert} (\mathbb{W}\widetilde{A}^T)_{i,j} \big{\rVert}_{2}^{2}
\end{equation}
Where $\mathbb{W}$ is the DFT matrix from eqn. (\ref{eqn-DFTMatrix}) and $\widetilde{A}^T$ is the transpose of the analysis frames matrix from eqn. (\ref{eqn-WindowMatrix}). The matrix $S$, is the spectrogram representation of the initial waveform, and has shape $N \times k$ of real floating-point numbers. We can index matrix $S$ similarly to that of matrix $A$ in eqn. (\ref{eqn-IndexingA}):
\begin{equation}
\label{eqn-IndexingS}
S_{i,j} = S[i][j] = S[i,j]
\end{equation}

@@

\paragraph*{}However, standard musical instruments seldom extend above $8$ kHz \cite{Olson,Virtanen,White}. This means that when constructing the spectrogram, we will rarely ever see energy present above this frequency at any time, and the $S$ matrix will contain mostly zero, or zero-like entries. As a result of this, we can simply crop each matrix to only represent frequencies that are only between $0$ Hz and $8000$ Hz. This makes the input array smaller, and eliminates redundant and non-useful information. The number rows in the $S$ matrix is reduced from $N$ down to $N'$. Since each raw audio file may contain a different number of samples to begin with, a different number of analysis frames are created. To ensure all samples are a homogeneous size, we assert that each matrix must have exactly $k'$ columns. If $k' < k$, columns of zeros are added to pad the input, and if $k' > k$, columns are removed.

@@

\paragraph*{}The features described in this section are derived from time-domain representations of each audio sample. Time space can be represented in one of three ways: 
\begin{itemize}
\item The raw waveform $s$\\
Indexed by $s[i]$ with $i \in [0,1,2,3,...,M-2,M-1]$
\item A single analysis frame $a_i$ \\
Indexed by $a_i[j]$ with $i \in [0,1,2,3,...,k-2,k-1]$ and $j \in [0,1,2,3,...,N-2,N-1]$
\item The full analysis frame matrix $A$ \\
Indexed by $A[i][j]$ with $i \in [0,1,2,3,...,k-2,k-1]$ and $j \in [0,1,2,3,...,N-2,N-1]$
\end{itemize}

\paragraph*{}From time space, we use the following $7$ features:
\begin{itemize}
\item[•] Time Domain Envelope
\item[•] Zero Crossing Rate
\item[•] Temporal Center of Mass
\item[•] Auto Correlation Coefficients ($\times 4$)
\end{itemize}

% ================================

@@

\paragraph*{}The time domain envelope (TDE) is a method of determining which parts of the sound wave contains most of the energy of the signal Often, we compute the RMS-Energy of each frame - a larger energy of the frame, the larger the average amplitude of the waveform in that frame. For frames with smaller amplitude, we can assume that the signal has either not begun, or is decaying.

\paragraph*{}We adapt this feature to \textbf{compute the RMS-Energy of the full waveform}. The RMS-Energy of the full waveform, $s$ is given by \cite{Olson,Virtanen}
\begin{equation}
\label{eqn-RMS}
\text{RMS}\big[ s \big] = \sqrt{\frac{1}{N} \sum_{i=0}^{N-1} s[i]^2}
\end{equation}
By using the full waveform, we acquire a rough estimate for the energy in the entirety of the audio file \cite{Liu}. For instruments with long sustain or release times, such as strings or undampened mallet percussion, we expect a comparably large waveform RMS when compared to those instruments without sustain such as plucked stings or dampened percussion

@@

\paragraph*{}The zero crossing rate (ZXR) of a signal or frame is use to measure how many times that a signal crosses it's equilibrium point. This can be done per total sound wave, per unit time, or per analysis-frame. This feature is most commonly associated with differentiating speech from music, because speech presents a more jagged and often less periodic waveform. In some cases, the ZXR can be correlated to frequency as well \cite{Kahn,Zhang}.

\paragraph*{}We adapt this feature to \textbf{compute the zero crossing rate for the full waveform}. The ZXR for the full waveform $s$ is given by \cite{Virtanen,Liu}
\begin{equation}
\label{eqn-ZXR}
\text{ZXR}\big[ s \big] = \frac{1}{2} \sum_{i=1}^{M-1} \big{|} \sign(s[i]) - \sign(s[i-1]) \big{|} 
\end{equation}
Where $\sign(x)$ returns $+1$ if $x > 0$, $-1$ if $x < 0$ and $0$ if $x = 0$. This provides a rough estimate for the average frequency in the full waveform, and can help discern clean periodic signals (low ZXR) from those that may have more noise or volatile behavior (high ZXR)\cite{Virtanen}.

@@

\paragraph*{}The temporal center of mass (TCM) of a signal is used to compute roughly where in time the amplitude of the waveform \textit{bunches up}. We treat the full waveform array, $s$, with $M$ samples as a 1-Dimensional discrete mass distribution. The TCM of that waveform is then given:
\begin{equation}
\label{eqn-FeatureTCM}
\text{TCM}\big[ s \big] = \frac{\sum_{i=0}^{M-1}i s[i]}{\sum_{i=0}^{M-1}s[i]}
\end{equation}
This includes negative amplitude values acting as "negative masses".

\paragraph*{}The TCM turns out to be a very unique feature, and very powerful in classification due to it's variance. For instruments with heavier attacks, we expect the TCM to be a very near the start of the waveform. For instruments with long sustain times, we expect a very centrally located center of mass, and for instruments with long release times, we expect a later TCM value.

@@

\paragraph*{}Auto correlation coefficients (ACC) are rough estimates of the signal spectral distribution . We can compute any number of ACC's and their value changed depending on the index chosen. It is common to only compute the first $K$ ACC's \cite{Virtanen}. For a full waveform signal $s$, with $M$ samples, the $k$-th ACC (indexed from $1$ to $K$) is given by:
\begin{equation}
\label{eqn-FeatureACC}
\text{ACC}_k\big[ s \big] = 
\frac{\sum_{i=0}^{M-k-1}s[i]s[i+k]} {\sqrt{\sum_{i=0}^{M-k-1}s^2[i]} \sqrt{\sum_{i=0}^{M-k-1}s^2[i+k]}}
\end{equation}

@@

\paragraph*{}The features described in this section are derived from the frequency-domain representations of each audio sample. Frequency space is represented by the transposed spectrogram matrix, $S^T$, described in sec. (\ref{subsec-spectrogram}). This is done to ensure that each  row is now an analysis frame, each column is a frequency bin. This gives a similar structure to the time-space analysis frames matrix, $A$, in eqn. (\ref{eqn-FrameMatrix}).
For each feature, we detail the physical significance and provide a visualization in feature-space.

\paragraph*{}From frequency space, we use the following $13$ features:
\begin{itemize}
\item[•] Mel Frequency Cepstral Coefficients ($\times 12$)
\item[•] Frequency Center of Mass
\end{itemize}

% ================================

@@

\paragraph*{}Mel Frequency Ceptstral Coefficients (MFCC's) are a very prolific feature used widly in audio classification tasks \cite{Liu,Virtanen,Zhang}. MFCC's are produced by converting a linear-frequency axis with units such as Hertz into units of Mels, which are design to account for the non-linear perception of frequency in humans. The Hertz to Mel and Mel to Hertz transforms are given \cite{Virtanen}:
\begin{equation}
\label{eqn-HztoMel}
M_f[h] = 2595 \log_{10}\big(1+ \frac{h}{700}\big)
\end{equation}
\begin{equation}
\label{eqn-MeltoHz}
H_f[m] = 700 \big(10^{(\frac{m}{2595})}-1\big)
\end{equation}
Where $M_f$ is the frequency in units of Mels, given $[h]$, a frequency in Hertz, and $H_f$ is the frequency in Hertz given $[m]$ a frequency in Mels.

@@

\paragraph*{}The frequency center-of-mass (FCM) for an audio file provides a strong representation of how overtones and energy is distributed in the signal's frequency domain. As with the temporal center-of-mass, we treat each row of the $S^T$ matrix as it's own 1-Dimensional mass distribution, and compute the center of mass of that row. This encodes the FCM for a single analysis frame (in frequency-space) in the waveform. For an frequency-analysis frame, $s^{(i)}$, the FCM is given by:
\begin{equation}
\label{eqn-FeatureFCM}
\text{TCM}_{i}\big[ s^{(i)} \big] = \frac{\sum_{j=0}^{N'-1}j s^{(i)}[j]}{\sum_{j=0}^{N'-1}s^{(i)}[j]}
\end{equation}
We compute the FCM for each of the $k'$ frames, and then average the results. We use the average FCM across $k'$ frames to compute the FCM feature:
\begin{equation}
TCM = \sum_{i=0}^{k'} \text{TCM}_{i}\big[ s^{(i)} \big]
\end{equation}
